Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-29 report/MSc_Report.tex
.

\chapter*{Abstract}
% Checklist (Abstract): [x] integrate supplied summary [ ] confirm PDF rebuild

This thesis studies multivariate density estimation for tabular data within a transport-based evaluation framework.
All models standardize features and learn a monotone lower-triangular map $S \colon \mathbf{u} \to \mathbf{z}$ to a Gaussian reference distribution.
The estimators accumulate exact log-Jacobians in standardized space with an affine correction for reporting on the data scale.
We compare separable triangular transport maps (TTM-Sep), additive-predictor Transformation Random Forests (TRTF), and copula baselines under the shared preprocessing pipeline.
Additive TRTF induces the same separable triangular likelihood as TTM-Sep via the probability integral transform.
The resulting estimator yields exact likelihoods, transparent conditionals, and linear per-sample cost but restricts context-dependent shape.

Benchmarks include synthetic Half-Moon and a four-dimensional autoregressive generator, plus the UCI POWER, GAS, HEPMASS, and MINIBOONE datasets.
For Half-Moon with $n = 250$, joint negative log-likelihoods (nats) were $1.71$ for TRTF, $1.93$ for TTM-Sep, and $1.54$ for the copula baseline.
For the four-dimensional generator, TRTF achieved $4.53$, TTM-Sep $5.66$, and the copula baseline $5.45$.
Permutation averages confirmed ordering sensitivity for triangular maps.
On MINIBOONE with $K = 43$ and $N = 2{,}500$, TRTF reached $-30.01$ test log-likelihood, whereas flow baselines remained near $-11.7$.

Findings indicate that TRTF narrows the gap to oracle likelihoods on low-dimensional synthetic data.
The method trails modern flows on high-dimensional real data because separability limits context-dependent shape.
Copula baselines remain competitive in low dimensions.
This thesis records guidance and limitations, including ordering effects and fragility in sparse regimes.

@@

We work in standardized coordinates to unify Jacobians, gradients, and reporting conventions. Let $x \in \mathbb{R}^K$ denote features on the original scale, and define
\begin{equation}
  u = T_{\text{std}}(x) = (x - \mu) \oslash \sigma,
  \qquad \sigma_k > 0,\label{eq:ch1-standardization}
\end{equation}
which standardizes each feature using training statistics only and fixes the coordinates where derivatives and Jacobians are evaluated. Let $S: u \mapsto z$ be a monotone lower-triangular map, and let $\eta$ denote the standard normal density. The change-of-variables identity gives
\begin{equation}
  \log \pi_U(u) = \log \eta\big(S(u)\big) + \log \big|\det \nabla_u S(u)\big|.
  \label{eq:ch1-pullback}
\end{equation}
Equation~\eqref{eq:ch1-pullback} splits the log density into a reference-fit term and a Jacobian term, isolating modeling capacity from the exact volume correction contributed by the map. For triangular $S$, the determinant factorizes as
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k} S_k(u_{1:k}).
  \label{eq:ch1-triangular-det}
\end{equation}
Equation~\eqref{eq:ch1-triangular-det} reduces the determinant to a sum of one-dimensional log derivatives, yielding linear per-sample complexity in $K$ and improving numerical stability. Reported log densities on the original scale apply the diagonal affine correction
\begin{equation}
  \log \pi_X(x) = \log \pi_U\!\big(T_{\text{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k.
  \label{eq:ch1-affine-correction}
\end{equation}
Equation~\eqref{eq:ch1-affine-correction} subtracts a constant offset determined by the training scales and ensures comparability across models that share standardized coordinates.
\medskip

We study separable triangular components for clarity, efficiency, and interpretability. Component $k$ decomposes into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),\qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k),
  \label{eq:ch1-separable}
\end{equation}
which stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.

% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
\section{Thesis and Problem Statement}\label{sec:ch1-problem}
This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame. We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests with axis-parallel splits (TRTF), and copula baselines.
\medskip

@@

We standardize features using training statistics only. Equation~\eqref{eq:ch1-standardization} defines the standardized coordinates $u$ used for evaluation. All derivatives and Jacobians are computed in $u$. The diagonal affine correction in Equation~\eqref{eq:ch1-affine-correction} reports log densities on the original scale $x$. This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets. All negative log-likelihoods are reported in nats.
\medskip

We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\phi$ and $\Phi$. Triangular transport maps are abbreviated TTM, with the separable variant labeled TTM-Sep. Transformation Random Forests are abbreviated TRTF, and the axis-parallel implementation is labeled TRTF. Copulas decouple marginals from dependence and serve as interpretable baselines.
\medskip

Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}. The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant. TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor. Copulas preserve explicit marginals and model dependence on the unit hypercube.

We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute. The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost. Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.

@@

The central problem is to determine when separability is appropriate for tabular data. We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention. Ordering effects, conditional calibration, and computational trade-offs address this question.
\medskip

On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MINIBOONE benchmark it improves on Gaussian references but trails published flow baselines. Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.

\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions. Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.

% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
\section{The Transport Frame on One Page}\label{sec:ch1-frame}
This section fixes the evaluation frame used across triangular transport maps, Transformation Random Forests, and copulas. It establishes notation, equations, and reporting conventions for the remainder of the thesis. The frame places all derivatives and Jacobians in standardized coordinates, applies a single diagonal affine correction on the original scale, and yields linear per-sample evaluation through a lower-triangular map.

Train-only standardization defines the evaluation coordinates via Equation~\eqref{eq:ch1-standardization}. The transformation uses training means and scales once and prevents information leakage into evaluation. All gradients and Jacobians act on $u$, which keeps objectives comparable across estimators and datasets. We convert to the original scale only when reporting.
\medskip

Let $S:u\mapsto z$ be a monotone lower-triangular map, and let $\eta(z)$ denote the $K$-variate standard normal density. The pullback identity in Equation~\eqref{eq:ch1-pullback} evaluates the reference at $S(u)$ and applies the exact volume correction. The split isolates model fit from geometry: the first term measures closeness to the Gaussian reference, and the second term contributes the exact Jacobian adjustment implied by the map.
\medskip

Lower-triangular structure with strictly positive diagonal partial derivatives factorizes the Jacobian as in Equation~\eqref{eq:ch1-triangular-det}. The factorization yields $\mathcal{O}(K)$ per-sample time for likelihood evaluation and improves numerical stability. It also guarantees global invertibility by back substitution consistent with the Rosenblatt and Knothe rearrangements.
\medskip

Reported log densities on the original scale apply only the diagonal affine correction implied by standardization. Equation~\eqref{eq:ch1-affine-correction} provides that adjustment. This convention keeps training and evaluation in $u$-space and converts to $x$-space at report time. All negative log-likelihoods are reported in nats, which fixes units across tables and figures.
\medskip

For clarity and efficiency we use separable triangular components of the form in Equation~\eqref{eq:ch1-separable}. Separable structure shifts location through $g_k$ and fixes conditional shape through the univariate monotone $h_k$. The Jacobian contribution depends only on $u_k$, which simplifies inversion by back substitution and stabilizes the log-determinant accumulation. The constraint also clarifies limits because conditional variance, skewness, and modality do not vary with context under separability.

The core operations follow the same path for all estimators.
\paragraph{Core operations.}
\begin{enumerate}
  \item First, standardize input features with training statistics using Equation~\eqref{eq:ch1-standardization}.
  \item Then, learn a monotone lower-triangular map $S$ to the standard normal reference and exploit Equation~\eqref{eq:ch1-pullback}.
  \item Finally, evaluate likelihoods, conditionals, and samples through Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction}.
\end{enumerate}

Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline. The input $x$ enters standardization to produce $u$. The triangular branch houses TTM-Sep and TRTF, whereas the copula branch couples fitted marginals to a dependence density. Both branches feed the same reported quantities, namely log density, conditionals, sampling, calibration, and compute. The appendix placement preserves the full schematic while keeping Chapter~\ref{ch:intro} focused on the narrative.

Notation remains consistent. We write $\eta$ for the $K$-variate standard normal density, and $\phi$ and $\Phi$ for the univariate standard normal density and CDF. We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$. These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.

@@

% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
\section{Contributions and Research Questions}\label{sec:ch1-contributions}
This section states the contributions of the thesis and formulates the research questions. It also maps both elements to the chapters and figures that deliver the evidence. The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow. All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.

The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines. The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map. It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators. Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.

The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting. We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data. The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation. Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.

@@

Two questions drive the empirical study and bind the contributions to specific measurements. The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data. All estimators share the transport frame in this comparison. We answer by reporting average test negative log-likelihoods in nats, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost. Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.

@@

@@

This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis. The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters. We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.

We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
\begin{equation}
  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
\end{equation}
where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division. In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.

The standardised density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$. Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$. The pullback identity then reads
\begin{equation}
  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
\end{equation}
which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant. Reporting log densities on the original scale requires only the diagonal affine correction implied by standardisation,
\begin{equation}
  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
\end{equation}
We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.

The transport is assumed to be lower triangular and componentwise monotone,
\begin{equation}
  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
\end{equation}
so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorises into a sum of one-dimensional log derivatives,
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
\end{equation}
The factorisation yields $\mathcal{O}(K)$ evaluation cost per sample, improves numerical stability, and guarantees global invertibility: strict monotonicity lets us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.

Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame. All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting. The remainder of this chapter adopts this frame. Section~\ref{sec:transport-separable} details the separable triangular parameterisation used for direct transports. Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform. Section~\ref{sec:transport-copula} places copulas in the same reporting convention.

\begin{table}[t]
  \centering
  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}. All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
  \label{tab:transport-notation}
  \begin{tabular}{ll}
    \hline
    Symbol & Meaning \\ 
    \hline
    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
    $T_{\mathrm{std}}$ & Standardisation map using training $(\mu,\sigma)$ \\ 
    $u = T_{\mathrm{std}}(x)$ & Standardised evaluation coordinates \\ 
    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
    $\eta(z)$ & $K$-variate standard normal density \\ 
    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
    $K$ & Dimension of the feature vector \\ 
    \hline
  \end{tabular}
\end{table}

@@

% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}. Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}. We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

@@

@@

The negative log-likelihood per sample takes the quadratic-plus-barrier form
\begin{equation}
  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
\end{equation}
which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out. Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier. In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

@@

In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized. The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape. Section~\ref{sec:transport-copula} positions copulas within the same reporting convention to decouple marginals from dependence.

@@

% Checklist: align references with transport frame, cite Sklar, enforce US spelling.
This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests. Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.

>>>>

\chapter*{Abstract}
% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild

This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF; additive predictor), and copulas (used only for low-dimensional diagnostics, $K\!\le\!3$). All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable. In the configuration studied (additive predictor and monotone CDF smoothing), TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.

On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula). On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps. On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings. These numbers are not strictly comparable but indicate the relative accuracy of this configuration.

Overall, TRTF tends to lead within the separable family at low dimension, while higher-dimensional datasets expose the limits of separable structure. We report robustness checks (ordering), calibration diagnostics, and the numerical safeguards used, and we outline directions toward richer parameterizations within the same evaluation frame.

@@

We adopt a shared evaluation protocol summarized in Section~\ref{sec:ch1-frame} and detailed in Chapter~\ref{ch:background}; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the pipeline at a glance.
\medskip

For clarity, efficiency, and interpretability we use separable triangular components as in Equation~\eqref{eq:transport-separable}. This structure stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.

% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
\section{Thesis and Problem Statement}\label{sec:ch1-problem}
This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame. We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copula baselines.
\medskip

@@

We standardize features using training statistics only. Equation~\eqref{eq:transport-standardise} defines the standardized coordinates $u$ used for evaluation. All derivatives and Jacobians are computed in $u$. The diagonal affine correction in Equation~\eqref{eq:transport-affine} reports log densities on the original scale $x$. This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets. All log quantities are reported in nats.
\medskip

We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\varphi$ and $\Phi$. Abbreviations for models and references appear in Table~\ref{tab:model-abbrev}. Copulas decouple marginals from dependence and serve as interpretable baselines.
\medskip

Separable triangular maps decompose each component into a context shift and a univariate monotone shape as in Equation~\eqref{eq:transport-separable}. The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant. Under strictly increasing conditional CDFs after standard monotone smoothing and with an additive predictor, TRTF implements the same separable triangular likelihood via the probability integral transform. Copulas preserve explicit marginals and model dependence on the unit hypercube; in this thesis they serve strictly as low-dimensional ($K\!\le\!3$) diagnostic baselines and are not evaluated on high-$K$ datasets.

We evaluate all estimators under the single protocol referenced above, with matched preprocessing and reporting to keep results comparable. Section~\ref{sec:evaluation-protocol} defines metrics and timing conventions.

@@

The central problem is to determine when separability is appropriate for tabular data. We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention (reported log densities on $x$ apply the affine correction in Eq.~\eqref{eq:transport-affine}). Ordering effects, conditional calibration, and computational trade-offs address this question.
\medskip

On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MiniBooNE benchmark it improves on Gaussian references but trails published flow baselines. Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.

\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions. We also exclude non-separable TRTF predictors and cross-term triangular maps because they raise compute and tuning costs substantially and complicate calibration/identifiability in our setting; the present scope focuses on separable structure for transparency and exact inversion. Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference. A brief pointer to expected changes if these variants were enabled—how cross-terms or non-additive predictors would likely affect calibration and NLL—is given in Chapter~\ref{ch:conclusion}, Section~\ref{sec:conclusion-outlook}.

% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
\section{The Transport Frame on One Page}\label{sec:ch1-frame}
To avoid duplication, the canonical derivations and notation live in Chapter~\ref{ch:background}. This section serves only as a map: we standardize with train-only statistics (Eq.~\eqref{eq:transport-standardise}), evaluate likelihoods via the pullback (Eq.~\eqref{eq:transport-pullback}), exploit the triangular determinant factorization (Eq.~\eqref{eq:transport-det}), and apply the affine correction for reporting (Eq.~\eqref{eq:transport-affine}). Separable components are defined in Eq.~\eqref{eq:transport-separable}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the pipeline.

Notation remains consistent. We write $\eta$ for the $K$-variate standard normal density, and $\varphi$ and $\Phi$ for the univariate standard normal density and CDF. We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$. These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.

@@

% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
\section{Contributions and Research Questions}\label{sec:ch1-contributions}
This section states the contributions and the research questions, and maps them to the chapters and figures that deliver the evidence. We adopt the shared transport frame summarized in Section~\ref{sec:ch1-frame}; Chapter~\ref{ch:background} records notation and assumptions, and Figure~\ref{fig:transport-schematic} anchors the comparisons.

The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests, and copula baselines. Where we claim an equivalence between TRTF and separable triangular transports, it holds under the conditions made explicit in Section~\ref{sec:transport-trtf} (strictly increasing conditional CDFs after monotone smoothing and an additive predictor). Chapter~\ref{ch:background} and Figure~\ref{fig:transport-schematic} establish the conventions and remove duplication in later chapters.

The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting. We evaluate TTM-Sep, TTM-Marg, and TRTF on synthetic generators and real tabular data; copulas are included only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not used in high-$K$ studies. The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation. Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.

@@

Two questions drive the empirical study and bind the contributions to specific measurements. The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data. All estimators share the transport frame in this comparison. We answer by reporting average test negative log-likelihoods, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost. Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.

@@

% Abbreviation table for reuse across chapters
\begin{table}[t]
  \centering
  \caption{Model abbreviations used throughout the thesis.}
  \label{tab:model-abbrev}
  \begin{tabular}{ll}
    \hline
    Label & Meaning \\
    \hline
    TTM-Marg & Marginal triangular transport map (per-dimension; no context) \\
    TTM-Sep & Separable triangular transport map (additive: $g_k$ shift + monotone $h_k$) \\
    TRTF & Transformation Random Forests (axis-parallel splits) \\
    True-Marg & Oracle marginal density \\
    True-Joint & Oracle conditional joint density \\
    Copula & Copula baseline (Gaussian or nonparametric) \\
    \hline
  \end{tabular}
\end{table}

@@

This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis. The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters. We summarize the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorization that drives evaluation and inversion.

We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
\begin{equation}
  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
\end{equation}
where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division. In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.

The standardized density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$. Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$. The pullback identity then reads
\begin{equation}
  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
\end{equation}
which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant. Reporting log densities on the original scale requires only the diagonal affine correction implied by standardization,
\begin{equation}
  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
\end{equation}
We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.

The transport is assumed to be lower triangular and componentwise monotone,
\begin{equation}
  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
\end{equation}
so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes into a sum of one-dimensional log derivatives,
\begin{equation}
  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
\end{equation}
The factorization yields $\mathcal{O}(K)$ evaluation cost per-sample, improves numerical stability, and guarantees global invertibility: strictly monotone diagonal derivatives let us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.

Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame. All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting. The remainder of this chapter adopts this frame. Section~\ref{sec:transport-separable} details the separable triangular parameterization used for direct transports. Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform under the conditions stated there (strictly increasing conditional CDFs after monotone smoothing and an additive predictor). Section~\ref{sec:transport-copula} places copulas in the same reporting convention.

\begin{table}[t]
  \centering
  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}. All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
  \label{tab:transport-notation}
  \begin{tabular}{ll}
    \hline
    Symbol & Meaning \\ 
    \hline
    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
    $\eta(z)$ & $K$-variate standard normal density \\ 
    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
    $K$ & Dimension of the feature vector \\ 
    \hline
  \end{tabular}
\end{table}

@@

% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}. Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$. The use of triangular transports builds on modern measure-transport literature; see, for instance, triangular transformations and their properties in \citet{bogachev2005triangular}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}. We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

@@

\begin{shaded}
\textbf{Assumptions.} Unless stated otherwise, we assume:
\begin{itemize}
  \item \emph{Lower-triangularity:} $S$ has the structure in Eq.~\eqref{eq:transport-triangular}.
  \item \emph{Strict monotone coordinates:} $\partial_{u_k} S_k(u_{1:k}) > 0$ for all $k$ and all arguments.
  \item \emph{Separable component:} Eq.~\eqref{eq:transport-separable} holds, so conditional shape along $u_k$ is fixed across contexts.
\end{itemize}
\end{shaded}

@@

The negative log-likelihood per-sample takes the quadratic-plus-barrier form
\begin{equation}
  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
\end{equation}
which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out. Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier. In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.

@@

In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized and the conditional CDFs are strictly increasing after monotone smoothing. The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape. Section~\ref{sec:transport-copula} positions copulas within the same reporting convention to decouple marginals from dependence.

@@

% Checklist: align references with transport frame, cite Sklar, enforce US spelling.
This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests. Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube \citep{nelsen2006introduction,joe2014dependence}. Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.
